\doctitle

### 2023-06-08 msr3.tex
Remove todos
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

% \edit{Rewrite the MSR3 abstract into a summary of whole work with citations}
\paragraph{Abstract for Chapter~\ref{ch:msr3}} Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
including cohort studies, longitudinal data analysis, and meta-analysis.  Design and analysis of variable selection methods for LMEs is more difficult than for linear regression because LME models are nonlinear. In this work we 
propose a relaxation strategy and optimization methods that enable a wide range of variable selection methods for LMEs using both convex and nonconvex regularizers, including  $\ell_1$, Adaptive-$\ell_1$, SCAD, and $\ell_0$. 
The computational framework only requires the proximal operator for each regularizer to be available, and the implementation is available in an open source \texttt{python} package \texttt{pysr3}, consistent 
with the \texttt{sklearn} standard. 
The numerical results on simulated data sets indicate that the proposed strategy improves on the state of the art for both accuracy and compute time. 
The variable selection techniques are also validated on a real example using a data set on bullying victimization.

@@

This extension and its mathematical foundations in \cite{Theory1} constitute

@@

foundations are presented in \cite{Theory1}.

@@

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{Theory1} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation

@@

Verification of the existence to solutions to \eqref{eq:lme_diagonal_setup}
and, more generally, \eqref{eq:ml_lme_optimization_setup} follows from the work of \cite{zheng2021trimmed}.
Standalone proofs for the existence of minimizers are developed in~\cite[Theorem 1]{Theory1},
and extended to the presence of regularizers in~\cite[Theorem 2]{Theory1}. %, restated below for convenience.

%%Here we streamline these techniques and extend to the regularized case, 
%%with basic results given below, with proofs in the appendix.  
%
%\begin{theorem}[Existence of a Minimizer]\label{thm:basic existence}
%Let the assumptions in the statement of problem \eqref{eq:ml_lme_optimization_setup} hold. Then optimal solutions to
%\eqref{eq:ml_lme_optimization_setup} exist.
%\end{theorem}
%
%Feature selection methods  include sparsity-inducing penalties or constraints. See~\cite[Theorem 1]{Theory1}
%for an existence result under weak assumptions on the regularizer. 
%
%
% 
%
%\begin{theorem}\label{thm:basic existence2}
%Let the assumptions in the statement of problem \eqref{eq:lme_diagonal_setup} hold,
%define $\LL(\beta,\gam):=\LL_{ML}(\beta, \Diag{\gamma})$
%and suppose $\map{\hR}{\R^p\times\R^q_+}{\R\cup\{+\infty\}}$ 
%is lsc and level bounded.
%Then $\LL+\hR$ is level bounded and solutions to
%the following optimization problem exist:
%\eq{
%\label{eq:extended loss}
%\min_{\beta\in\R^p, \gamma\in\R^q_+}  \LL(\beta, \gamma) + 
%\hR(\beta, \gamma) .
%}
%\end{theorem}

@@

%Feature selection is imposed by adding
%Consider a mixed-effect model setting described in Eq. (\ref{eq:lme_setup}) with $\Gamma$ being diagonal: $\Gamma = \Diag{\gamma}$.  
%We wish to find a minimizer of the problem 
%\eq{
%	\label{eq:lme_loss_original}
%	\min_{\beta\in\R^p, \gamma\in\R^q_+} & \LL(\beta, \gamma) + R(\beta, \gamma), 
%	}
%\vskip -.5in
\eq{%\label{eq:main opt in x over C}
     \min_{x } & \LL(x) + R(x)+ \delta_{\CC}(x),
      \label{eq:lme_loss_original_in_x}
%    \\
%    \text{s.t. } & x \in \CC\ .
}
where $x = (\beta, \gamma)$, 
$\CC:=\R^p\times \R^q_+$,
$\map{R}{\R^P\times\R^q_+}{\eR_+:=\R_+\cup\{+\infty\}}$
is a 
lower semi-continuous (lsc) regularization term, and
$\delta_{\CC}$ is the convex indicator function, where
$\delta_{\CC}(x) :=0$ for $x \in \CC$ and $+\infty$ otherwise.
%\[
%    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
%\]  
By \cite[Theorem 2]{Theory1}, solutions to \eqref{eq:lme_loss_original_in_x}
always exist when $R$ has compact lower level sets.
% or 
% $\LL(\beta, \gamma)=\LL_{REML}(\beta,\Diag(\gamma))$ in \eqref{eq:reml_objective} with 
% \[
% \B=[\zero,\gam_\mmax\one]:=\left\{ \gamma\, |\, 0\le \gamma_i\le \gam_\mmax,\ i=1,\dots,q\right\}
% \]
% for some $\gam_\mmax>0$ where $\one$ is the vector of the appropriate
% dimension each of whose components is one.
% Since the method applies to either likelihood, hereafter we omit the subscripts and use $\LL$ to denote the objective function. %, leaving this choice up to a practitioner.
%For conciseness, define $x = (\beta, \gamma)$ and 
%$\CC:=\R^p\times \R^q_+$.
%such that $\tbeta \in \R^p$ is unconstrained and $\tgamma \in \R^q_+$ is non-negative:
%\eq{
% x\in \CC \textabove{$\Delta$}{=} \{[\beta', \gamma']: \  \beta' \in \R^p, \ \gamma' \in \R^q_+\} \subset \R^{p+q} 
%}
%With this notation, problem \eqref{eq:lme_loss_original} becomes
%\eq{%\label{eq:main opt in x over C}
%    \label{eq:lme_loss_original_in_x}
%    \min_{x \in \CC} & \LL(x) + R(x),
%    \\
%    \text{s.t. } & x \in \CC\ .
%}
%Equivalently, the box constraint can be moved to the loss in a form of an indicator function:
%or equivalently
%\[
%\eq{
%	\label{eq:lme_loss_original}
%    \min_{x} \LL(x) + R(x) + \delta_{\CC}(x)\ ,
%}
%where 
%$x = (\beta, \gamma)$, 
%$\CC:=\R^p\times \R^q_+$,  
%\quad \text{where}\ \, 
%\delta_{\CC}$ is the convex indicator function
%\[
%    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
%\]
The most common regularizers are separable taking the form
\begin{equation}\label{eq:R sep}
R(x) = \sum_{i = 1}^p r_i(x_i), 
\end{equation}
with typical choices for the component functions $r_i$ given in Table \ref{table:proxes}.

@@

with $\eta\ge\bar\eta$ where $\bar \eta$ is the weak convexity constant
computed in~\cite[Section 5.1]{Theory1}.
As $\eta \uparrow \infty$, the extended objective~\eqref{eq:lme_loss_sr3} converges in an epigraphical sense to the original objective~\eqref{eq:lme_loss_original_in_x}. 
However, feature selection accuracy does not require this continuation, 
indeed, we show that a fixed 
modest value such as $\eta=1$ can be used~\citep{Zheng2019SR3}.

@@

In the case of linear regression, 
the function $u_\eta$ has a closed form solution~\cite{Zheng2019SR3}. However, in both  the linear regression context of~\cite{Zheng2019SR3} 
and in the LME context studied here, we need only compute $S_\eta(w)$ 
in order to optimize~\eqref{eq:value_function_optimization}. 
Indeed, in \cite[Section 5]{Theory1} it is shown that 
there exists a computable $\bar\eta>0$, which we have called the weak convexity constant, such that
$\LL+\delta_\CC+\kappa_\eta(\cdot -w)$ is strongly convex for all 
$\eta>\bar\eta$ regardless of the choice of $w$. This allows us to show
that 
$u_\eta$ is well-defined, 
differentiable, and Lipschitz continuous, with
%and we have a closed form solution for the derivative: 
\begin{equation}\label{eq:grad_value_function_explicit}
\nabla u_\eta(w) = \nabla_w k_\eta(x-w)|_{x = S_\eta(w)}=\eta(w-S_\eta(w)) .  
\end{equation}
%We call $\bar\eta$ the weak convexity constant for 
%$\LL+\delta_\CC+\kappa_\eta(\cdot -w)$.
%where $\nabla_w k_\eta(x-w)$ is computed directly from~\eqref{eq:kappa}.

@@

We apply PGD to optimize the regularized value function $u_\eta$ which yields the iteration 
\eq{
	\label{eq:pgd_for_value_function}
	w^+ = \prox_{\alpha^{-1}R}(w - \alpha\nabla u_\eta(w))
}
The results in \cite{Theory1} show that all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
The equivalence of Algorithm \ref{alg:pgd_for_value} and 
\eqref{eq:pgd_for_value_function} is established in the following lemma, 
which extends the relationship studied by~\cite{Zheng2019SR3} to the case of $x = (\beta, \gamma)$.

@@

In \cite[Theorem 6]{Theory1}, it is shown that
for any sequence $\eta_k\uparrow\infty$ the associated optimal solutions $(x^k,w^k)$ 
to (11) satisfy
\(
\mathcal{L}(x^k)+R(w^k)\uparrow 
\inf_{x\in\Rp\times\Rq_+}\mathcal{L}(x)+R(x) \;
\text{with}\; \Vert{x^k-w^k}\Vert\rightarrow 0.
\)
In particular, every cluster point of the sequences $\{x^k\}$ and $\{w^k\}$ are solutions
to \eqref{eq:lme_diagonal_setup}, where such cluster points exist whenever
the function $R$ is coercive, i.e. $\lim_{\norm{x}\uparrow\infty}R(x)=+\infty$.
Just how close $w^k$ is to a solution to \eqref{eq:lme_diagonal_setup} remains
an open question, however, our numerical studies in Section \ref{sec:applications}
show that $\eta$ can be chosen surprisingly small. Indeed, we typically take $\eta=1$.

@@

\paragraph{Positive Approximation of the Hessian}
For many datasets the weak convexity constant $\overline \eta$ can be extremely large 
and difficult to compute. However, if $ \eta$ is too small 
$\nabla^2_{\gamma\gamma}\LL_{\mu,\eta}(\beta, \gamma)$ is 
negative-(semi)definite. Negative definite Hessians can hamper the convergence of 
second-order methods (e.g., see \cite{nocedal2006numerical}). 
Therefore, one must take care in selecting $\eta$. For this, we recall from
\cite[Lemma 3]{Theory1} that
\[
\nabla^2\LL{(\beta,\gam)}=\sum_{i=1}^m
S_i^T\begin{bmatrix}X_i^T\\ -Z_i^T\end{bmatrix}
\Omega_i(\gam)^{-1}
\begin{bmatrix}X_i& -Z_i\end{bmatrix}S_i
-\begin{bmatrix}
0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
\end{bmatrix}.
\]
This implies that negative eigenvalues for the Hessian must arise from the
Hessian with respect to $\gamma$,
$\nabla^2_{\gamma\gamma}\LL(\beta, \gamma)$, and more specifically, the
term $(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}$. 
A positive semidefinite approximation to the Hessian is obtained by simply dropping this term.

@@

\section{Theoretical Analysis}
\label{sec:theory-paper}
\input{msr3-theory}

\section{Software Implementation}
\label{sec:software-paper}
To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library~(\cite{sholokhov2023pysr3}). This library implements functionality for fitting linear mixed models and selecting covariates. The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{sklearn} library to minimize learning time.

>>>>

\paragraph{Abstract for Chapter~\ref{ch:msr3}} Linear Mixed-Effects (LME) models are a fundamental tool for modeling correlated data, 
including cohort studies, longitudinal data analysis, and meta-analysis.  Design and analysis of variable selection methods for LMEs is more difficult than for linear regression because LME models are nonlinear. In this work we 
propose a relaxation strategy and optimization methods that enable a wide range of variable selection methods for LMEs using both convex and nonconvex regularizers, including  $\ell_1$, Adaptive-$\ell_1$, SCAD, and $\ell_0$. 
The computational framework only requires the proximal operator for each regularizer to be available, and the implementation is available in an open source \texttt{python} package \texttt{pysr3}, consistent 
with the \texttt{scikit-learn} standard. 
The numerical results on simulated data sets indicate that the proposed strategy improves on the state of the art for both accuracy and compute time. 
The variable selection techniques are also validated on a real example using a data set on bullying victimization.

@@

This extension and its mathematical foundations in \cite{aravkin2022jimtheory} constitute

@@

foundations are presented in \cite{aravkin2022jimtheory}.

@@

The negative log likelihood~\eqref{eq:ml_lme_optimization_setup}  is nonlinear and nonconvex, and requires an iterative numerical solver.
However, it is convex with respect to $\beta$, and weakly convex with respect to $\gamma$, with a weak convexity constant $\overline \eta$
computed in~\cite[Section 5.1]{aravkin2022jimtheory} . The expected value of the posterior mode $\beta$ given $\Gamma$ has the closed form representation

@@

Verification of the existence to solutions to \eqref{eq:lme_diagonal_setup}
and, more generally, \eqref{eq:ml_lme_optimization_setup} follows from the work of \cite{zheng2021trimmed}.
Standalone proofs for the existence of minimizers are developed in~\cite[Theorem 1]{aravkin2022jimtheory},
and extended to the presence of regularizers in~\cite[Theorem 2]{aravkin2022jimtheory}. %, restated below for convenience.

%%Here we streamline these techniques and extend to the regularized case, 
%%with basic results given below, with proofs in the appendix.  
%
%\begin{theorem}[Existence of a Minimizer]\label{thm:basic existence}
%Let the assumptions in the statement of problem \eqref{eq:ml_lme_optimization_setup} hold. Then optimal solutions to
%\eqref{eq:ml_lme_optimization_setup} exist.
%\end{theorem}
%
%Feature selection methods  include sparsity-inducing penalties or constraints. See~\cite[Theorem 1]{aravkin2022jimtheory}
%for an existence result under weak assumptions on the regularizer. 
%
%
% 
%
%\begin{theorem}\label{thm:basic existence2}
%Let the assumptions in the statement of problem \eqref{eq:lme_diagonal_setup} hold,
%define $\LL(\beta,\gam):=\LL_{ML}(\beta, \Diag{\gamma})$
%and suppose $\map{\hR}{\R^p\times\R^q_+}{\R\cup\{+\infty\}}$ 
%is lsc and level bounded.
%Then $\LL+\hR$ is level bounded and solutions to
%the following optimization problem exist:
%\eq{
%\label{eq:extended loss}
%\min_{\beta\in\R^p, \gamma\in\R^q_+}  \LL(\beta, \gamma) + 
%\hR(\beta, \gamma) .
%}
%\end{theorem}

@@

%Feature selection is imposed by adding
%Consider a mixed-effect model setting described in Eq. (\ref{eq:lme_setup}) with $\Gamma$ being diagonal: $\Gamma = \Diag{\gamma}$.  
%We wish to find a minimizer of the problem 
%\eq{
%	\label{eq:lme_loss_original}
%	\min_{\beta\in\R^p, \gamma\in\R^q_+} & \LL(\beta, \gamma) + R(\beta, \gamma), 
%	}
%\vskip -.5in
\eq{%\label{eq:main opt in x over C}
     \min_{x } & \LL(x) + R(x)+ \delta_{\CC}(x),
      \label{eq:lme_loss_original_in_x}
%    \\
%    \text{s.t. } & x \in \CC\ .
}
where $x = (\beta, \gamma)$, 
$\CC:=\R^p\times \R^q_+$,
$\map{R}{\R^P\times\R^q_+}{\eR_+:=\R_+\cup\{+\infty\}}$
is a 
lower semi-continuous (lsc) regularization term, and
$\delta_{\CC}$ is the convex indicator function, where
$\delta_{\CC}(x) :=0$ for $x \in \CC$ and $+\infty$ otherwise.
%\[
%    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
%\]  
By \cite[Theorem 2]{aravkin2022jimtheory}, solutions to \eqref{eq:lme_loss_original_in_x}
always exist when $R$ has compact lower level sets.
% or 
% $\LL(\beta, \gamma)=\LL_{REML}(\beta,\Diag(\gamma))$ in \eqref{eq:reml_objective} with 
% \[
% \B=[\zero,\gam_\mmax\one]:=\left\{ \gamma\, |\, 0\le \gamma_i\le \gam_\mmax,\ i=1,\dots,q\right\}
% \]
% for some $\gam_\mmax>0$ where $\one$ is the vector of the appropriate
% dimension each of whose components is one.
% Since the method applies to either likelihood, hereafter we omit the subscripts and use $\LL$ to denote the objective function. %, leaving this choice up to a practitioner.
%For conciseness, define $x = (\beta, \gamma)$ and 
%$\CC:=\R^p\times \R^q_+$.
%such that $\tbeta \in \R^p$ is unconstrained and $\tgamma \in \R^q_+$ is non-negative:
%\eq{
% x\in \CC \textabove{$\Delta$}{=} \{[\beta', \gamma']: \  \beta' \in \R^p, \ \gamma' \in \R^q_+\} \subset \R^{p+q} 
%}
%With this notation, problem \eqref{eq:lme_loss_original} becomes
%\eq{%\label{eq:main opt in x over C}
%    \label{eq:lme_loss_original_in_x}
%    \min_{x \in \CC} & \LL(x) + R(x),
%    \\
%    \text{s.t. } & x \in \CC\ .
%}
%Equivalently, the box constraint can be moved to the loss in a form of an indicator function:
%or equivalently
%\[
%\eq{
%	\label{eq:lme_loss_original}
%    \min_{x} \LL(x) + R(x) + \delta_{\CC}(x)\ ,
%}
%where 
%$x = (\beta, \gamma)$, 
%$\CC:=\R^p\times \R^q_+$,  
%\quad \text{where}\ \, 
%\delta_{\CC}$ is the convex indicator function
%\[
%    \delta_{\CC}(x) := \begin{cases} 0, &  x \in \CC \\ +\infty, & x \not\in \CC .\end{cases}
%\]
The most common regularizers are separable taking the form
\begin{equation}\label{eq:R sep}
R(x) = \sum_{i = 1}^p r_i(x_i), 
\end{equation}
with typical choices for the component functions $r_i$ given in Table \ref{table:proxes}.

@@

with $\eta\ge\bar\eta$ where $\bar \eta$ is the weak convexity constant
computed in~\cite[Section 5.1]{aravkin2022jimtheory}.
As $\eta \uparrow \infty$, the extended objective~\eqref{eq:lme_loss_sr3} converges in an epigraphical sense to the original objective~\eqref{eq:lme_loss_original_in_x}. 
However, feature selection accuracy does not require this continuation, 
indeed, we show that a fixed 
modest value such as $\eta=1$ can be used~\citep{Zheng2019SR3}.

@@

In the case of linear regression, 
the function $u_\eta$ has a closed form solution~\cite{Zheng2019SR3}. However, in both  the linear regression context of~\cite{Zheng2019SR3} 
and in the LME context studied here, we need only compute $S_\eta(w)$ 
in order to optimize~\eqref{eq:value_function_optimization}. 
Indeed, in \cite[Section 5]{aravkin2022jimtheory} it is shown that 
there exists a computable $\bar\eta>0$, which we have called the weak convexity constant, such that
$\LL+\delta_\CC+\kappa_\eta(\cdot -w)$ is strongly convex for all 
$\eta>\bar\eta$ regardless of the choice of $w$. This allows us to show
that 
$u_\eta$ is well-defined, 
differentiable, and Lipschitz continuous, with
%and we have a closed form solution for the derivative: 
\begin{equation}\label{eq:grad_value_function_explicit}
\nabla u_\eta(w) = \nabla_w k_\eta(x-w)|_{x = S_\eta(w)}=\eta(w-S_\eta(w)) .  
\end{equation}
%We call $\bar\eta$ the weak convexity constant for 
%$\LL+\delta_\CC+\kappa_\eta(\cdot -w)$.
%where $\nabla_w k_\eta(x-w)$ is computed directly from~\eqref{eq:kappa}.

@@

We apply PGD to optimize the regularized value function $u_\eta$ which yields the iteration 
\eq{
	\label{eq:pgd_for_value_function}
	w^+ = \prox_{\alpha^{-1}R}(w - \alpha\nabla u_\eta(w))
}
The results in \cite{aravkin2022jimtheory} show that all components of the iteration~(\ref{eq:pgd_for_value_function}) are well-defined. 
The equivalence of Algorithm \ref{alg:pgd_for_value} and 
\eqref{eq:pgd_for_value_function} is established in the following lemma, 
which extends the relationship studied by~\cite{Zheng2019SR3} to the case of $x = (\beta, \gamma)$.

@@

In \cite[Theorem 6]{aravkin2022jimtheory}, it is shown that
for any sequence $\eta_k\uparrow\infty$ the associated optimal solutions $(x^k,w^k)$ 
to (11) satisfy
\(
\mathcal{L}(x^k)+R(w^k)\uparrow 
\inf_{x\in\Rp\times\Rq_+}\mathcal{L}(x)+R(x) \;
\text{with}\; \Vert{x^k-w^k}\Vert\rightarrow 0.
\)
In particular, every cluster point of the sequences $\{x^k\}$ and $\{w^k\}$ are solutions
to \eqref{eq:lme_diagonal_setup}, where such cluster points exist whenever
the function $R$ is coercive, i.e. $\lim_{\norm{x}\uparrow\infty}R(x)=+\infty$.
Just how close $w^k$ is to a solution to \eqref{eq:lme_diagonal_setup} remains
an open question, however, our numerical studies in Section \ref{sec:applications}
show that $\eta$ can be chosen surprisingly small. Indeed, we typically take $\eta=1$.

@@

\paragraph{Positive Approximation of the Hessian}
For many datasets the weak convexity constant $\overline \eta$ can be extremely large 
and difficult to compute. However, if $ \eta$ is too small 
$\nabla^2_{\gamma\gamma}\LL_{\mu,\eta}(\beta, \gamma)$ is 
negative-(semi)definite. Negative definite Hessians can hamper the convergence of 
second-order methods (e.g., see \cite{nocedal2006numerical}). 
Therefore, one must take care in selecting $\eta$. For this, we recall from
\cite[Lemma 3]{aravkin2022jimtheory} that
\[
\nabla^2\LL{(\beta,\gam)}=\sum_{i=1}^m
S_i^T\begin{bmatrix}X_i^T\\ -Z_i^T\end{bmatrix}
\Omega_i(\gam)^{-1}
\begin{bmatrix}X_i& -Z_i\end{bmatrix}S_i
-\begin{bmatrix}
0&0\\ 0& \half(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}
\end{bmatrix}.
\]
This implies that negative eigenvalues for the Hessian must arise from the
Hessian with respect to $\gamma$,
$\nabla^2_{\gamma\gamma}\LL(\beta, \gamma)$, and more specifically, the
term $(Z_i^T\Omega_i(\gam)^{-1}Z_i)^{\circ2}$. 
A positive semidefinite approximation to the Hessian is obtained by simply dropping this term.

@@

\newpage
\section{Theoretical Analysis}
\label{sec:theory-paper}
\input{msr3-theory}

\newpage
\section{Software Implementation}
\label{sec:software-paper}
To ensure reproducibility of this research, all new algorithms have been implemented as a part of the \texttt{pysr3}\footnote{Available at \href{https://github.com/aksholokhov/pysr3}{https://github.com/aksholokhov/pysr3}} library~(\cite{sholokhov2023pysr3}). This library implements functionality for fitting linear mixed models and selecting covariates. The user interface was designed to be fully compliant with the standards\footnote{\href{https://scikit-learn.org/stable/developers/develop.html}{https://scikit-learn.org/stable/developers/develop.html}} of \texttt{scikit-learn} library to minimize learning time.

### 2023-06-08 phd_thesis.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

\begin{document}

\title{\doctitle}
~\\
{\center
{\large 
{\large
\projectNameLong
}
\\[10mm]
\userName
\\[20mm]
{A dissertation \\
 submitted in partial fulfillment of the\\ requirements for the degree of}
\\[5mm]
{Doctor of Philosophy}
\\[20mm]
{\center
\textbf{Reading Committee}\\
Aleksandr Aravkin (co-chair),\\
J. Nathan Kutz (co-chair), \\
James Burke, \\
Jonathan Wakefield (GSR)
}
\\[20mm]
Program Authorized to Offer Degree:
\\[5mm]
Applied Mathematics
\\[5mm]
\institution
}
\\[10mm]
©Copyright 2023\\
\userName

@@

\newpage
{
\center 
\institution
\\[10mm]
\textbf{Abstract}
\\[10mm]
\projectNameLong
\\[10mm]
\userName
\\[10mm]
Chairs of the Supervisory Committee:\\
Aleksandr Aravkin and J. Nathan Kutz\\
\department
\\[10mm]
}

This work summarizes two projects focused on incorporating prior knowledge into machine learning models. In the first project, a universal feature selection method for linear mixed-effect models is developed. Namely, Sparse Relaxed Regularized Regression (SR3) is extended to the case of Linear Mixed-Effect (LME) likelihoods, and we prove that one can minimize such likelihoods with proximal gradient descent. Theoretical underpinnings of the proposed extension are also presented, including consistency results, variational properties, implementability of optimization methods, and convergence results. In particular, convergence analyses are provided for a basic implementation of SR3 for LME and an accelerated hybrid algorithm. Numerical results show the utility and speed of these algorithms on real and simulated datasets. Finally, both algorithms are implemented in an open-source python package \texttt{pysr3}. Conveniently, this package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in a pipeline with classic modeling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics.

The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE). In particular, a classic technique of collocation points is adapted to transfer knowledge from a known equation to a model that approximates solutions of that equation. The addition of a physics-informed loss allows for exceptional data supply strategies that improve the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. The resulting ROMs extrapolate forward in time more accurately, perform better for unseen initial conditions, and exhibit less sensitivity to noise. Finally, I show how such ROMs can be used as strong regularizers in single-pixel imaging (SPI), enabling the reduction of samples-per-frame rate by an order of magnitude relative to other state-of-the-art algorithms.

\tableofcontents

\chapter{Introduction}
\include{introduction}

@@

\chapter{PINODE: Physics-Informed Neural Ordinary Differential Equations}
\label{ch:pinode}
\include{pinode}

@@

\chapter{Conclusion}
\include{conclusion}

@@

\appendix
\chapter{Appendix}
\section{Acknowledgement}
\input{appendix/acknowledgements}

@@

\section{Detailed Design of Experiments for PINODE}
\input{appendix/pinode_appendix}

>>>>

\begin{document}
\thispagestyle{empty}
\title{\doctitle}
~\\
{\center
{\large 
{\large
\projectNameLong
}
\\[10mm]
\userName
\\[15mm]
{A dissertation \\
 submitted in partial fulfillment of the\\ requirements for the degree of}
\\[10mm]
{Doctor of Philosophy}
\\[10mm]
\institution
\\[5mm]
2023
\\[15mm]
{\center
Reading Committee:\\
Aleksandr Aravkin (chair),\\
Jose Kutz (chair), \\
James Burke
}
\\[15mm]
Program Authorized to Offer Degree:
\\[5mm]
Applied Mathematics
\\[5mm]
}
}

\newpage
\thispagestyle{empty}
{
\center

©Copyright 2023\\
\userName

@@

\newpage
\thispagestyle{empty}
{
\center 
\institution
\\[10mm]
\textbf{Abstract}
\\[10mm]
\projectNameLong
\\[10mm]
\userName
\\[10mm]
Chairs of the Supervisory Committee:\\
Aleksandr Aravkin and J. Nathan Kutz\\
\department
\\[10mm]
}

This work summarizes two projects focused on incorporating prior knowledge into machine learning models. In the first project, a universal feature selection method for linear mixed-effect models is developed. Namely, Sparse Relaxed Regularized Regression (SR3) is extended to the case of Linear Mixed-Effect (LME) likelihoods, and we prove that one can minimize such likelihoods with proximal gradient descent. Theoretical underpinnings of the proposed extension are also presented, including consistency results, variational properties, implementability of optimization methods, and convergence results. In particular, convergence analyses are provided for a basic implementation of SR3 for LME and an accelerated hybrid algorithm. Numerical results show the utility and speed of these algorithms on real and simulated datasets. Finally, both algorithms are implemented in an open-source python package \texttt{pysr3}. Conveniently, this package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in a pipeline with classic modeling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics. The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE). In particular, a classic technique of collocation points is adapted to transfer knowledge from a known equation to a model that approximates solutions of that equation. The addition of a physics-informed loss allows for exceptional data supply strategies that improve the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. The resulting ROMs extrapolate forward in time more accurately, perform better for unseen initial conditions, and exhibit less sensitivity to noise. Finally, I show how such ROMs can be used as strong regularizers in single-pixel imaging (SPI), enabling the reduction of samples-per-frame rate by an order of magnitude relative to other state-of-the-art algorithms.

\newpage
\tableofcontents

\chapter*{Introduction}
\include{introduction}

@@

\addtocontents{toc}{\protect\newpage}
\chapter{PINODE: Physics-Informed Neural Ordinary Differential Equations}
\label{ch:pinode}
\include{pinode}

@@

\chapter*{Conclusion}
\include{conclusion}

@@

\addtocontents{toc}{\protect\newpage}

\appendix
\chapter{MSR3}

\section*{Acknowledgement}
\input{appendix/acknowledgements}

@@

\chapter{PINODE}
\section{Detailed Design of Experiments for PINODE}
\input{appendix/pinode_appendix}
